\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2021-07-05 doc/USER_MANUAL/02_getting_started.tex

To download the SPECFEM3D\_Cartesian software package, type this:
\begin{verbatim}
git clone --recursive --branch devel https://github.com/geodynamics/specfem3d.git
\end{verbatim}
%
\noindent
\red{Note: for people who would like to run the package on Windows rather than on Unix machines, you can install Docker or VirtualBox (installing a Linux in VirtualBox in that latter case) and run it easily from inside that.}\\

\noindent
We recommend that you add {\texttt{ulimit -S -s unlimited}} to your
{\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize
unlimited }} to your {\texttt{.cshrc}} file to suppress any potential
limit to the size of the Unix stack.\\

@@

You can replace the GNU compilers above (gfortran and gcc) with other compilers if you want to; for instance for Intel ifort and icc use FC=ifort CC=icc instead.

Note that MPI must be installed with MPI-IO enabled because parts of SPECFEM3D perform I/Os through MPI-IO.

Before running the \texttt{configure} script, you should probably
edit file \texttt{flags.guess} to make sure that it contains the best
compiler options for your system. Known issues or things to check
are:
\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine. \textbf{In the case of that compiler, we have noticed that initial release versions sometimes have bugs or issues that can lead to wrong results when running the code, thus we \emph{strongly} recommend using a version for which at least one service pack or update has been installed.}
\red{In particular, for version 17 of that compiler, users have reported problems (making the code crash at run time) with the \texttt{-assume buffered\_io} option; if you notice problems,
remove that option from file \texttt{flags.guess} or change it to \texttt{-assume nobuffered\_io} and try again.}
\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave}
or \texttt{-qnosave} for your machine.
\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
\end{description}
When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc}
compilers, we suggest running the \texttt{configure} script with the
following options:

@@

If you have problems configuring the code on a Cray machine, i.e. for instance if you get an error message from the \texttt{configure} script, try exporting these two variables:
\texttt{MPI\_INC=\${CRAY\_MPICH2\_DIR}/include and FCLIBS=" "}, and for more details if needed you can refer to the \texttt{utils/Cray\_compiler\_information} directory.
You can also have a look at the configure script called \texttt{utils/Cray\_compiler\_information/configure\_SPECFEM\_for\_Piz\_Daint.bash}.

@@

You can add \texttt{--enable-vectorization} to the configuration options to speed up the code in the fluid (acoustic) and elastic parts.
This works fine if (and only if) your computer always allocates a contiguous memory block for each allocatable array;
this is the case for most machines and most compilers, but not all. To disable this feature, use option \texttt{--disable-vectorization}.
For more details see \href{https://github.com/geodynamics/specfem3d/issues/81}{github.com/geodynamics/specfem3d/issues/81} .
To check if that option works fine on your machine, run the code with and without it for an acoustic/elastic model and make sure the seismograms are identical.\\

Note that we use CUBIT (now called Trelis) to create meshes of hexahedra, but other packages
can be used as well, for instance GiD from \url{http://gid.cimne.upc.es}
or Gmsh from \url{http://geuz.org/gmsh} \citep{GeRe09}. Even mesh
creation packages that generate tetrahedra, for instance TetGen from
\url{http://tetgen.berlios.de}, can be used because each tetrahedron
can then easily be decomposed into four hexahedra as shown in the
picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif};
while this approach does not generate hexahedra of optimal quality,
it can ease mesh creation in some situations and it has been shown
that the spectral-element method can very accurately handle distorted
mesh elements \citep{OlSe11}.

@@

\noindent
When compiling for GPU cards, you can use CUDA with:
\begin{verbatim}
  ./configure --with-cuda ..
\end{verbatim}
or
\begin{verbatim}
  ./configure --with-cuda=cuda4 ..
\end{verbatim}
where for example \texttt{cuda4,cuda5,cuda6,cuda7,..} specifies the target GPU architecture of your card,
(e.g., with CUDA 4 this refers to Tesla K10 cards), rather than the installed version of the CUDA toolkit.
Before CUDA version 5, one version supported basically one new architecture and needed a different kind of compilation.
Since version 5, the compilation has stayed the same, but newer versions supported newer architectures.
However at the moment, we still have one version linked to one specific architecture:
\begin{verbatim}
- CUDA 4 for Tesla,   cards like K10, Geforce GTX 650, ..
- CUDA 5 for Kepler,  like K20
- CUDA 6 for Kepler,  like K80
- CUDA 7 for Maxwell, like Quadro K2200
- CUDA 8 for Pascal,  like P100
- CUDA 9 for Volta,   like V100
- CUDA 10 for Turing, like GeForce RTX 2080
- CUDA 11 for Ampere, like A100
\end{verbatim}
So even if you have the new CUDA toolkit version 11, but you want to run on say a K20 GPU, then you would still configure with:
\begin{verbatim}
  ./configure --with-cuda=cuda5
\end{verbatim}
The compilation with the cuda5 setting chooses then the right architecture (\texttt{-gencode=arch=compute\_35,code=sm\_35} for K20 cards).

@@

A summary of the most important configuration variables follows.
\begin{description}
\item [{\texttt{F90}}] Path to the Fortran compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flags.
\end{description}
The configuration script automatically creates for each executable
a corresponding \texttt{Makefile} in the \texttt{src/} subdirectory.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{constants.h} file depending on your
system:
\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database
for the solver on the local disks of the compute nodes. Some systems
have no local disks, e.g., BlueGene or the Earth Simulator, and other
systems have a fast parallel file system, in which case this flag
should be set to \texttt{.true.}. Note that this flag is not used
by the database generator or the solver; it is only used for some
of the post-processing.
\end{description}
The package can run either in single or in double precision mode.
The default is single precision because for almost all calculations
performed using the spectral-element method using single precision
is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half
as much memory. Select your preference by selecting the appropriate
setting in the \texttt{constants.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
In the \texttt{precision.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
On many current processors (e.g., Intel, AMD, IBM Power), single precision
calculations are significantly faster; the difference can typically
be 10\% to 25\%. It is therefore better to use single precision. What
you can do once for the physical problem you want to study is run
the same calculation in single precision and in double precision on
your system and compare the seismograms. If they are identical (and
in most cases they will), you can select single precision for your
future runs.\\

@@

\section{Adding OpenMP support in addition to MPI}

NOTE FROM JULY 2013: OpenMP support is maybe / probably not maintained any more.
Thus the section below is maybe obsolete.

OpenMP support can be enabled in addition to MPI. However, in many
cases performance will not improve because our pure MPI implementation
is already heavily optimized and thus the resulting code will in fact
be slightly slower. A possible exception could be IBM BlueGene-type
architectures.

To enable OpenMP, uncomment the OpenMP compiler option in two lines
in file \texttt{src/specfem3D/Makefile.in} (before running \texttt{configure})
and also uncomment the
\texttt{\#define USE\_OPENMP} statement in file\\
\texttt{src/specfem3D/specfem3D.F90}.

\section{Compiling on an IBM BlueGene}

\underline{More recent installation instruction for IBM BlueGene, from April 2013:}\\

@@

\section{Visualizing the subroutine calling tree of the source code}

@@

\section{Using the ADIOS library for I/O}

Regular POSIX I/O can be problematic when dealing with large simulations one large
clusters (typically more than $10,000$ processes). SPECFEM3D use the ADIOS library~\cite{Liu2013}
to deal transparently take advantage of advanced parallel file system features. To enable
ADIOS, the following steps should be done:
\begin{enumerate}
\item Install ADIOS (available from \url{https://www.olcf.ornl.gov/center-projects/adios/}).
Make sure that your environment variables reference it.
\item You may want to change ADIOS related values in the \texttt{constants.h.in} file.
The default values probably suit most cases.
\item Configure using the \texttt{--with-adios} flag.
\end{enumerate}
ADIOS is currently only usable for meshfem3D generated mesh (i.e. not for meshes generated
with CUBTI). Additional control parameters are discussed in section~\ref{cha:Main-Parameter}.

\section{Becoming a developer of the code, or making small modifications in the source code}

If you want to develop new features in the code, and/or if you want to make small changes, improvements, or bug fixes, you are very welcome to contribute. To do so, i.e. to access the development branch of the source code with read/write access (in a safe way, no need to worry too much about breaking the package, there is a robot called BuildBot that is in charge of checking and validating all new contributions and changes), please visit this Web page: \url{https://github.com/geodynamics/specfem3d/wiki/Using-Hub}.

>>>>

To download the SPECFEM3D\_Cartesian software package, type this:
\begin{verbatim}
git clone --recursive --branch devel https://github.com/geodynamics/specfem3d.git
\end{verbatim}
%

@@

You can replace the GNU compilers above (gfortran and gcc) with other compilers if you want to; for instance for Intel ifort and icc use FC=ifort CC=icc instead. Note that MPI must be installed with MPI-IO enabled because parts of SPECFEM3D perform I/Os through MPI-IO.

Before running the \texttt{configure} script, you should probably
edit file \texttt{flags.guess} to make sure that it contains the best
compiler options for your system. Known issues or things to check
are:
\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine. In the case of that compiler, we have noticed that initial release versions sometimes have bugs or issues that can lead to wrong results when running the code, thus we \emph{strongly} recommend using a version for which at least one service pack or update has been installed.
In particular, for version 17 of that compiler, users have reported problems (making the code crash at run time) with the \texttt{-assume buffered\_io} option; if you notice problems,
remove that option from file \texttt{flags.guess} or change it to \texttt{-assume nobuffered\_io} and try again.

\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave}
or \texttt{-qnosave} for your machine.

\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
\end{description}

When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc}
compilers, we suggest running the \texttt{configure} script with the
following options:

@@

If you have problems configuring the code on a Cray machine, i.e. for instance if you get an error message from the \texttt{configure} script, try exporting these two variables:
\texttt{MPI\_INC=\${CRAY\_MPICH2\_DIR}/include and FCLIBS=" "}, and for more details if needed you can refer to the \texttt{utils/Cray\_compiler\_information} directory.
You can also have a look at the configure script called:\\
\texttt{utils/Cray\_compiler\_information/configure\_SPECFEM\_for\_Piz\_Daint.bash}.

@@

You can add \texttt{-{}-enable-vectorization} to the configuration options to speed up the code in the fluid (acoustic) and elastic parts.
This works fine if (and only if) your computer always allocates a contiguous memory block for each allocatable array;
this is the case for most machines and most compilers, but not all. To disable this feature, use option \texttt{-{}-disable-vectorization}.
For more details see \href{https://github.com/geodynamics/specfem3d/issues/81}{github.com/geodynamics/specfem3d/issues/81} .
To check if that option works fine on your machine, run the code with and without it for an acoustic/elastic model and make sure the seismograms are identical.\\

Note that we use CUBIT (now called Trelis) to create meshes of hexahedra, but other packages
can be used as well, for instance GiD from \url{http://gid.cimne.upc.es}
or Gmsh from \url{http://geuz.org/gmsh} \citep{GeRe09}. Even mesh
creation packages that generate tetrahedra, for instance TetGen from
\url{http://tetgen.berlios.de}, can be used because each tetrahedron
can then easily be decomposed into four hexahedra as shown in the
picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif};
while this approach does not generate hexahedra of optimal quality,
it can ease mesh creation in some situations and it has been shown
that the spectral-element method can very accurately handle distorted
mesh elements \citep{OlSe11}.\\

@@

@@

For people who would like to run the package on Windows rather than on Unix machines, you can install Docker or VirtualBox (installing a Linux in VirtualBox in that latter case) and run it easily from inside that.\\

We recommend that you add {\texttt{ulimit -S -s unlimited}} to your
{\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize
unlimited }} to your {\texttt{.cshrc}} file to suppress any potential
limit to the size of the Unix stack.\\

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Using the GPU version of the code}
%-----------------------------------------------------------------------------------------------------------------------------------%

\noindent
SPECFEM3D now supports CUDA and HIP GPU acceleration.
When compiling for GPU cards, you can enable the CUDA version with:
\begin{verbatim}
  ./configure --with-cuda ..
\end{verbatim}
or
\begin{verbatim}
  ./configure --with-cuda=cuda9 ..
\end{verbatim}
where for example \texttt{cuda4,cuda5,cuda6,cuda7,..} specifies the target GPU architecture of your card,
(e.g., with CUDA 9 this refers to Volta V100 cards), rather than the installed version of the CUDA toolkit.
Before CUDA version 5, one version supported basically one new architecture and needed a different kind of compilation.
Since version 5, the compilation has stayed the same, but newer versions supported newer architectures.
However at the moment, we still have one version linked to one specific architecture:
\begin{verbatim}
- CUDA 4 for Tesla,   cards like K10, Geforce GTX 650, ..
- CUDA 5 for Kepler,  like K20
- CUDA 6 for Kepler,  like K80
- CUDA 7 for Maxwell, like Quadro K2200
- CUDA 8 for Pascal,  like P100
- CUDA 9 for Volta,   like V100
- CUDA 10 for Turing, like GeForce RTX 2080
- CUDA 11 for Ampere, like A100
\end{verbatim}
So even if you have the new CUDA toolkit version 11, but you want to run on say a K20 GPU, then you would still configure with:
\begin{verbatim}
  ./configure --with-cuda=cuda5
\end{verbatim}
The compilation with the cuda5 setting chooses then the right architecture (\texttt{-gencode=arch=compute\_35,code=sm\_35} for K20 cards).\\

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Using the ADIOS library for I/O}
%-----------------------------------------------------------------------------------------------------------------------------------%

Regular POSIX I/O can be problematic when dealing with large simulations on large
clusters (typically more than $10,000$ MPI processes). SPECFEM3D can use the ADIOS library~\cite{Liu2013}
to take advantage of advanced parallel file system features. To enable
ADIOS, the following steps should be done:
\begin{enumerate}
\item Install ADIOS (available from \url{https://www.olcf.ornl.gov/center-projects/adios/}).
Make sure that your environment variables reference it.
\item You may want to change ADIOS related values in the \texttt{constants.h.in} file.
The default values probably suit most cases.
\item Configure using the \texttt{-{}-with-adios} flag.
\end{enumerate}
ADIOS is currently only usable for meshfem3D generated mesh (i.e. not for meshes generated
with CUBIT). Additional control parameters are discussed in section~\ref{cha:Main-Parameter}.

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Adding OpenMP support in addition to MPI}
%-----------------------------------------------------------------------------------------------------------------------------------%

OpenMP support can be enabled in addition to MPI. However, in many
cases performance will not improve because our pure MPI implementation
is already heavily optimized and thus the resulting code will in fact
be slightly slower. A possible exception could be IBM BlueGene-type
architectures.\\

\noindent
To enable OpenMP,  add the flag \texttt{-{}-enable-openmp} to the configuration:
\begin{verbatim}
./configure --enable-openmp ..
\end{verbatim}
This will add the corresponding OpenMP flag for the chosen Fortran compiler.\\

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Configuration summary}
%-----------------------------------------------------------------------------------------------------------------------------------%

\noindent
A summary of the most important configuration variables follows.
\begin{description}
\item [{\texttt{F90}}] Path to the Fortran compiler.

\item [{\texttt{MPIF90}}] Path to MPI Fortran.

\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.

\item [{\texttt{FLAGS\_CHECK}}] Compiler flags.
\end{description}
%
The configuration script automatically creates for each executable
a corresponding \texttt{Makefile} in the \texttt{src/} subdirectory.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.\\

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{constants.h} file depending on your
system:
\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database
for the solver on the local disks of the compute nodes. Some systems
have no local disks, e.g., BlueGene or the Earth Simulator, and other
systems have a fast parallel file system, in which case this flag
should be set to \texttt{.true.}. Note that this flag is not used
by the database generator or the solver; it is only used for some
of the post-processing.
\end{description}
%
The package can run either in single or in double precision mode.
The default is single precision because for almost all calculations
performed using the spectral-element method using single precision
is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half
as much memory. Select your preference by selecting the appropriate
setting in the \texttt{constants.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
%
In the \texttt{precision.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
%
On many current processors (e.g., Intel, AMD, IBM Power), single precision
calculations are significantly faster; the difference can typically
be 10\% to 25\%. It is therefore better to use single precision. What
you can do once for the physical problem you want to study is run
the same calculation in single precision and in double precision on
your system and compare the seismograms. If they are identical (and
in most cases they will), you can select single precision for your
future runs.\\

@@

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Compiling on an IBM BlueGene}
%-----------------------------------------------------------------------------------------------------------------------------------%

\noindent
Installation instructions for IBM BlueGene (from April 2013):\\

@@

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Visualizing the subroutine calling tree of the source code}
%-----------------------------------------------------------------------------------------------------------------------------------%

@@

%-----------------------------------------------------------------------------------------------------------------------------------%
\section{Becoming a developer of the code, or making small modifications in the source code}
%-----------------------------------------------------------------------------------------------------------------------------------%

If you want to develop new features in the code, and/or if you want to make small changes, improvements, or bug fixes, you are very welcome to contribute. To do so, i.e. to access the development branch of the source code with read/write access (in a safe way, no need to worry too much about breaking the package, there are CI tests based on BuildBot, Travis-CI and Jenkins in place that are checking and validating all new contributions and changes), please visit this Web page:\\
\url{https://github.com/geodynamics/specfem3d/wiki}

### 2021-07-05 doc/USER_MANUAL/08b_performing_full_waveform_inversion_FWI_or_source_inversions.tex

\red{\Large ANNOUNCEMENT: SPECFEM3D can now perform full waveform inversion (FWI), i.e. invert for models in an iterative fashion, and it can also perform source inversions in a constant model;
please refer to the two new directories, \texttt{inverse\_problem\_for\_model} and \texttt{inverse\_problem\_for\_source}, and the README files they contain. For FWI inversions for the model, also refer to the new examples provided in
directory \texttt{EXAMPLES}.}

>>>>

SPECFEM3D can now perform full waveform inversion (FWI), i.e. invert for models in an iterative fashion, and it can also perform source inversions in a constant model;
please refer to the two new directories, \texttt{inverse\_problem\_for\_model} and \texttt{inverse\_problem\_for\_source}, and the README files they contain. For FWI inversions for the model, also refer to the new examples provided in
directory \texttt{EXAMPLES}.
